### 2021-07-09 Nonlinear/formats2021/introduction.tex
Submitted version of formats2021 paper

One of the widely used techniques for performing safety analysis of nonlinear dynamical systems is to perform reachable set computation.
%
Reachable set is an overapproximation of the set of states visited by all the trajectories of the system starting from an initial set.
%
Computing the reachable set for nonlinear systems is very challenging.
%
This is primarily because of two reasons:
%
First, the tools for performing nonlinear analysis are not very scalable.
%
Second, computing the reachable set using set representations involves wrapping error.
%
That is, the overapproximation acquired at a given step would increase the conservativeness of the reachable set for all future steps.

One of the techniques for computing reachable set for discrete time nonlinear systems is to use Parallelotope bundles.
%
Here, the reachable set is represented as a parallelotope bundle, an intersection of several parallelotopes.
%
One of the advantages of this technique is that it uses a special form of nonlinear optimization problem to compute the reachable set.
%
The usage of a specific form of nonlinear optimization mitigates the drawback involved with the scalability of nonlinear analysis.

However, wrapping error still remains to be a problem for reachability using parallelotope bundles.
%
The template directions for specifying these parallelotopes are provided as an input by the user.
%
Often, these template directions are selected to be either the cardinal axis directions or some directions from octahedral domains.
%
However, it is not clear that the axis directions and octagonal directions are optimal for computing reachable sets.
%
Also, even an expert user of reachable set computation tools might not be able to provide the most suitable set of template directions for computing the most accurate reachable set.
%
Picking an unsuitable template directions would only cause the wrapping error to increase, thus increasing the conservativeness of the safety analysis.

In this paper, we investigate techniques for generating template directions automatically and dynamically.
%
That is, instead of providing the template directions to compute the parallelotope, the user just specifies the number of templates and the algorithm automatically generates the template directions.
%
We study two techniques for generating the template directions.
%
First, we compute a local linear approximation of the nonlinear dynamics and use the linear approximation to compute the templates.
%
Second, we generate a specific set of sample trajectories from the set and use principal component analysis (PCA) over these trajectories.
%
We observe that the accuracy of the reachable set can be drastically improved by using templates generated using these two techniques.
%
For standard nonlinear benchmark systems, we show that generating templates in a dynamic fashion improves the accuracy of the reachable set by two orders of magnitude.
%
We demonstrate that even when the size of the initial set increases, our template generation technique returns much more accurate reachable sets than both manually specified and random template directions.

@@

Reachable set computation of nonlinear systems using template polyhedra and Bernstein polynomials has been first proposed in~\cite{dang2009image}.
%
In~\cite{dang2009image}, Bernstein polynomial representation is used to compute an upper bound of a special type of nonlinear optimization problem~\cite{garloff2003bernstein}.
%
Several improvements to this algorithm were suggested in~\cite{dang2012reachability,sassi2021reachability} and~\cite{dang2014parameter} extends it for performing parameter synthesis.
%
The representation of parallelotope bundles for reachability was proposed in~\cite{dreossi2016parallelotope} and the effectiveness of using bundles for reachability was demonstrated in~\cite{dreossi2017sapo,dreossi2017reachability}.
%
However, all of these papers used static template directions for computing the reachable set.

Using template directions for reachable set has been proposed in~\cite{sankaranarayanan2008symbolic} and later improved in~\cite{dang2011template}.
%
Leveraging the principle component analysis of sample trajectories for computing reachable set has been proposed in~\cite{stursberg2003efficient,chen2011choice,seladji2017finding}.
%
More recently, connections between optimal template directions for reachability of linear dynamical systems and bilinear programming have been highlighted in~\cite{gronski2019template}.
%
For static template directions, Octahedral domain directions~\cite{clariso2004octahedron} remain a popular choice.

>>>>

One of the most widely-used techniques for performing safety analysis of nonlinear dynamical systems is reachable set computation.
%
The reachable set is defined to be the set of states visited by at least one of the trajectories of the system starting from an initial set.
%
Computing the reachable set for nonlinear systems is challenging primarily due to two reasons:
%
First, the tools for performing nonlinear analysis are not very scalable.
%
Second, computing the reachable set using set representations involves wrapping error.
%
That is, the overapproximation acquired at a given step would increase the conservativeness of the overapproximation for all future steps.

One of the techniques for computing the overapproximation of reachable sets for discrete time nonlinear systems is to use parallelotope bundles.
%
Here, the reachable set is represented as a parallelotope bundle, an intersection of several parallelotopes.
%
One of the advantages of this technique is its utilization of a special form of nonlinear optimization problem to overapproximate the reachable set.
%
The usage of a specific form of nonlinear optimization mitigates the drawback involved with the scalability of nonlinear analysis.

However, wrapping error still remains to be a problem for reachability using parallelotope bundles.
%
The template directions for specifying these parallelotopes are provided as an input by the user.
%
Often, these template directions are selected to be either the cardinal axis directions or some directions from octahedral domains.
%
However, it is not clear that the axis directions and octagonal directions are optimal for computing reachable sets.
%
Also, even an expert user of reachable set computation tools may not be able to provide a suitable set of template directions for computing reasonably accurate over-approximations of the reachable set.
%
Picking unsuitable template directions would only cause the wrapping error to increase, thus increasing the conservativeness of the safety analysis.

In this paper, we investigate techniques for generating template directions automatically and dynamically.
%
That is, instead of providing the template directions to compute the parallelotope, the user just specifies the number of templates and the algorithm automatically generates the template directions.
%
We study two techniques for generating the template directions.
%
First, we compute a local linear approximation of the nonlinear dynamics and use the linear approximation to compute the templates.
%
Second, we generate a specific set of sample trajectories from the set and use principal component analysis (PCA) over these trajectories.
%
We observe that the accuracy of the reachable set can be drastically improved by using templates generated using these two techniques.
%
For standard nonlinear benchmark systems, we show that generating templates in a dynamic fashion improves the accuracy of the reachable set by two orders of magnitude.
%
We demonstrate that even when the size of the initial set increases, our template generation technique returns more accurate reachable sets than both manually-specified and random template directions.

@@

Reachable set computation of nonlinear systems using template polyhedra and Bernstein polynomials has been first proposed in~\cite{dang2009image}.
%
In~\cite{dang2009image}, Bernstein polynomial representation is used to compute an upper bound of a special type of nonlinear optimization problem~\cite{garloff2003bernstein}.
%
Several improvements to this algorithm were suggested in~\cite{dang2012reachability,sassi2012reachability} and~\cite{dang2014parameter} extends it for performing parameter synthesis.
%
The representation of parallelotope bundles for reachability was proposed in~\cite{dreossi2016parallelotope} and the effectiveness of using bundles for reachability was demonstrated in~\cite{dreossi2017sapo,dreossi2017reachability}.
%
However, all of these papers used static template directions for computing the reachable set.

Using template directions for reachable set has been proposed in~\cite{sankaranarayanan2008symbolic} and later improved in~\cite{dang2011template}.
%
Leveraging the principal component analysis of sample trajectories for computing reachable set has been proposed in~\cite{stursberg2003efficient,chen2011choice,seladji2017finding}.
%
More recently, connections between optimal template directions for reachability of linear dynamical systems and bilinear programming have been highlighted in~\cite{gronski2019template}.
%
For static template directions, octahedral domain directions~\cite{clariso2004octahedron} remain a popular choice.

### 2021-07-09 Nonlinear/formats2021/prelims.tex
Submitted version of formats2021 paper

A parallelotope $P$, denoted as a tuple $\tup{a, G}$ where $a \in \reals^n$ is called the \emph{anchor} and $G$ is a set of vectors $\{g_1, g_2, \ldots, g_n\}$, $\forall_{1 \leq i \leq n} ~ g_i \in \reals^n$ called \emph{generators}, represents a set
\begin{equation}
P = \{ x\:|\: \exists \alpha_1, \ldots, \alpha_n, \mbox{ such that }  0 \leq \alpha_i \leq 1, x = a + \sum_{i=1}^n \alpha_i g_i \}.
\label{eq:ptope}
\end{equation}

@@

  To compute the generator representation of $P$, we need to compute the \emph{anchor} and the \emph{generators}.
  %
  The anchor is obtained by solving the linear equations $x-y = 0, y = 0$.
  %
  Therefore, the anchor $a$ is the vertex at origin $(0,0)$
  %
  To compute the two generator of the parallelotope, we compute two vertices of the parallelotope.
  %
  Vertex $v_1$ is obtained by solving the linear equations $x - y = 1, y = 0$.
  %
  Therefore, vertex $v_1$ is the vertex $(1,0)$.
  %
  Similarly, vertex $v_2$ is obtained by solving the linear equations $x-y = 0, y = 1$.
  %
  Therefore, $v_2$ is the vertex $(1,1)$.
  %
  The generator $g_1$ is the vector $v_1 - a$, that is $(1,0)- (0,0) = (1,0)$
  %
  The generator $g_2$ is the vector $v_2 - a$, that is $(1,1) - (0,0) = (1,1)$.
  %
  Therefore, all the points in the paralellotope can be written as $(x,y) = (0,0) + \alpha_1 (1,0) + \alpha_2(1,1)$, $0 \leq \alpha_1, \alpha_2 \leq 1$.
  %
  % We can also convert a generator representation of a parallelotope into the half-plane representation.
  %
\end{example}

A parallelotope bundle $Q$ is a set of parallelotopes $\{P_1, \ldots, P_m\}$.
%
The set of states represented by a parallelotope bundle is given as the intersection
\begin{equation}
Q = \bigcap_{i=1}^m P_i.
\end{equation}
%
Often, the various parallelotopes in a bundle share common template directions.
%
In such cases, the conjunction of all the parallelotope constraints in a bundle $Q$ is written as $c_{l}^Q \leq T^Q x \leq c_{u}^Q$.
%
Notice that the number of upper and lower bound half-space constraints in this bundle are stricly more than $n$, i.e., $T^Q \in R^{m \times n}$ where $m>n$.
%
Each parallelotope in such a bundle is represented as a subset of constraints in $c_{l}^Q \leq T^Q x \leq c_{u}^Q$.
%
These types of bundles are often considered in the literature.

Alternatively, in this paper, we consider parallelotope bundles where the consisting parallelotopes do not share template directions.
%
We consider such bundles because our we generate template directions automatically.
%
We propose techniques to generate $n$ template directions at each instance.
%
% collectively generate one set of template directions.

@@

For computing the reachable set of a nonlinear system, we need an upper bound for the optimization problem.
%
Several techniques using interval arithmetic and Bernstein polynomials have been developed in the recent past~\cite{garloff2003bernstein,munoz2013formalization,kodiak,smith2009fast}.
%

>>>>

A parallelotope $P$, denoted as a tuple $\tup{a, G}$ where $a \in \reals^n$ is called the \emph{anchor} and $G$ is a set of vectors $\{g_1, g_2, \ldots, g_n\}$, $\forall_{1 \leq i \leq n} ~ g_i \in \reals^n$ called \emph{generators}, represents the set
\begin{equation}
P = \{ x\:|\: \exists \alpha_1, \ldots, \alpha_n, \mbox{ such that }  0 \leq \alpha_i \leq 1, x = a + \sum_{i=1}^n \alpha_i g_i \}.
\label{eq:ptope}
\end{equation}

@@

  To compute the generator representation of $P$, we need to compute the \emph{anchor} and the \emph{generators}.
  %
  The anchor is obtained by solving the linear equations $x-y = 0, y = 0$.
  %
  Therefore, the anchor $a$ is the vertex at origin $(0,0)$
  %
  To compute the two generators of the parallelotope, we compute two vertices of the parallelotope.
  %
  Vertex $v_1$ is obtained by solving the linear equations $x - y = 1, y = 0$.
  %
  Therefore, vertex $v_1$ is the vertex $(1,0)$.
  %
  Similarly, vertex $v_2$ is obtained by solving the linear equations $x-y = 0, y = 1$.
  %
  Therefore, $v_2$ is the vertex $(1,1)$.
  %
  The generator $g_1$ is the vector $v_1 - a$, that is $(1,0)- (0,0) = (1,0)$
  %
  The generator $g_2$ is the vector $v_2 - a$, that is $(1,1) - (0,0) = (1,1)$.
  %
  Therefore, all the points in the paralellotope can be written as $(x,y) = (0,0) + \alpha_1 (1,0) + \alpha_2(1,1)$, $0 \leq \alpha_1, \alpha_2 \leq 1$.
  %
  % We can also convert a generator representation of a parallelotope into the half-plane representation.
  %
\end{example}

A parallelotope bundle $Q$ is a set of parallelotopes $\{P_1, \ldots, P_m\}$.
%
The set of states represented by a parallelotope bundle is given as the intersection
\begin{equation}
Q = \bigcap_{i=1}^m P_i.
\end{equation}
%
Often, the various parallelotopes in a bundle share common template directions.
%
In such cases, the conjunction of all the parallelotope constraints in a bundle $Q$ is written as $c_{l}^Q \leq T^Q x \leq c_{u}^Q$.
%
Notice that the number of upper and lower bound half-space constraints in this bundle are stricly more than $n$ in these cases, i.e., $T^Q \in R^{m \times n}$ where $m>n$.
%
Each parallelotope in such a bundle is represented as a subset of constraints in $c_{l}^Q \leq T^Q x \leq c_{u}^Q$.
%
These types of bundles are often considered in the literature. \cite{dang2009image,dang2012reachability,dreossi2017reachability}

Alternatively, we consider parallelotope bundles where the consisting parallelotopes do not share template directions.
%
We consider such bundles because we generate the $n$ template directions automatically at each step.
%
% collectively generate one set of template directions.

@@

For computing the reachable set of a nonlinear system, we need an upper bound for the optimization problem.
%
Several techniques using interval arithmetic and Bernstein polynomials have been developed in the recent past \newline\cite{kodiak,garloff2003bernstein,munoz2013formalization,smith2009fast}.
%

### 2021-07-09 Nonlinear/formats2021/theory.tex
Submitted version of formats2021 paper

\section{Reachability Algorithm}
\label{sec:theory}

In this work, we develop parallelotope reachability algorithms that are \textbf{automatic} with \textbf{dynamic} parallelotopes.
%
The state of the art, in contrast, is \textbf{manual}, where the user specifies a set of parallelotope directions at the start of the problem.
%
Further, the parallelotopes are \textbf{static}, and do not change during the course of the computation.
%
In our evaluation in the next section, we will demonstrate the benefits of these change on a series of nonlinear example systems.
%
In this section, we detail the modifications to the algorithm present correctness arguments.

@@

We first present the original algorithm where the user manually specifies the number of parallelotopes and a set of directions for each parallelotope that remain static throughout the computation.

Recall the system is $n$-dimensional with dynamics function $f: \reals^{n} \rightarrow \reals^{n}$.
%
The parallelotope bundle $Q$ is specified as a collection of $m$ template directions $\T^{Q} \in \reals^{m \times n} (m > n)$ and the set of constraints that define each of the consisting parallelotope.
%
% consisting of $m$ parallelotopes , the user should manually provide $n$ generator directions for each parallelotope.
%
% We write the collective set of generator directions as $\mathcal{G}$, which is a set of sets, each of which contains $n$ direction vectors.
%
Another input to the algorithm is the initial set, given as a parallelotope $P_0$.
%
When the initial set is a box, $P_0$ consists has axis-aligned template directions.
%
The output of the algorithm is, for each step $k$, the set $\overline\Theta_k$, which is a conservative overapproximation of the reachable set at step $k$, $\Theta_k \subseteq \overline\Theta_k$.

The high-level pseudo-code is written in Algorithm~\ref{alg:old}.
%
The algorithm simply calls \tbundle for each step, producing a new parallelotope bundle based on the parallelotope bundle from the previous step.
%
To compute the image of $Q$, the algorithm computes the upper and lower bounds of $f(x)$ with respect to each template direction.
%
Since computing the maximum value of $f(x)$ along each template direction on $Q$ is computationally very hard, the algorithm instead computes the maximum value over each of the consisting parallelotopes and uses the mimimum of all these maximum values.
%
The \tbundle operation works as follows.
%
Consider a parallelotope $P$ in the bundle $Q$.
%
From the definition, it follows that $Q \subseteq P$.
%
% and half-space representation as $\tup{\T, c_{l}, c_{u}$.
Given a template direction $\T_i$, the maximum value of $\T_{i} f(x)$ for all $x \in Q$ is less than or equal to the maximum value of $\T_{i} f(x)$ for all $x \in P$.
%
Similar argument holds for the mimum value of $T_{i} f(x)$ for all $x \in Q$.
%
% We would like to compute the image of the parallelotope $P$ with respect to the nonlinear transformation $f(x)$.
%
% That is, we would like to compute $P'$ such that $\forall x \in P, f(x) \in P'$.
%
% Given that current techniques use static template directions, the parallelotope $P'$ also uses the template directions $\T^{P}$.
%
To compute the upper and lower bounds of each template direction $\T_{i} f(x)$, for all $x \in P$, we perform the following optimization.
%
\begin{eqnarray}
  \max ~ \T_i^{P} \cdot f(x) \label{eq:maxf}\\
  s.t. ~~ x \in P.\nonumber
\end{eqnarray}

Given $P$ is a parallelotope, all the states in $P$ can be expressed as a vector summation of anchor and scaled generators.
%
Let  $\tup{a, G}$ be the generator representation of $P$.
%
The optimization problem given in Equation~\ref{eq:maxf} would then transform as follows.

@@

Equation~\ref{eq:maxalpha} is a form of $\mathsf{optBox(\T_{i} \cdot f)}$ over $[0,1]^n$.
%
For computing an upper-bound to the constrained nonlinear optimization by invoking one of the Bernstein polynomial or interval arithmetic based methods.
%
Similarly, we compute the lowerbound of $\T_{i}f(x)$ for all $x \in P$ by computing the upperbound of $-1 \times \T_{i}f(x)$.
%

We iterate this process (i.e., computing the upper and lower bound of $T_{i}f(x)$) for each parallelotope in the bundle $Q$.
%
% Observe that the role of the parallelotopes in the bundle is to compute an upper bound on the maximum value of $T_{i}f(x)$ by expanding the domain from $Q$ to the corresponding parallelotope $P$.
%
Therefore, the tightest upper bound on $T_{i}f(x)$ over $Q$ is the least of the upper bounds computed from each of the parallelotopes.
%
Similar argument holds for lower bounds of $T_{i}f(x)$ over $Q$.
%
Therefore, the image of the bundle $Q$ will be the bundle $Q'$ where the upper and lower bounds for templates directions are obtained by solving several constrained nonlinear optimization problems.

@@

The proposed automatic dynamic algorithm does not require the user provide the set of template direction $\T$; instead it creates templates automatically at each step based.
%
We use two techniques to generate such template directions, first: computing local linear approximations of the dynamics and second, perform principle component analysis (PCA) over sample trajectories.
%
To do this, we first sample a set of points in the parallelotope bundle called \emph{support points}, and propagate them to the next step using the dynamics function $f$.
%
Support points are a subset of the vertices of the parallelotope that either maximize or minimize the template directions.

Intuitively, linear approximations can provide good approximations when the dynamics function is a time-discretization of a continuous system.
%
In this case, for small time steps a nonlinear function can be approximated fairly accurately by a linear function.
%
We use the support points as a data-driven approach to find the best-fit linear function to use.
%
If the dynamics of a system is linear, i.e., $x^{+} = Ax$, the image of the parallelotope $c_{l} \leq \T x \leq c_{u}$, is the set $c_{l} \leq \T \cdot A^{-1} x \leq c_{u}$.
%
Therefore, given the template directions of the initial set as $\T_0$, we compute the local linear approximation of the nonlinear dynamics (from the support points) and change the template directions by multiplying it with the inverse of the approximate linear dynamics.
%
The second technique for generating template directions performs principle component analysis over the images of the support points.
%
Using PCA is a reasonable choice as it produces orthonormal directions that can construct a rotated box for bounding the points.
%

Observe that in general, the dynamics is nonlinear and therefore, the reachable set could non-convex.
%
However, a parallelotope bundle is always a convex set.
%
Therefore, one way to improve accuracy of this representation is to consider more template directions.
%
For this purpose, we use a notion of \emph{template lifespan}, where we use the linear approximation and/or PCA template directions not only from the current step, but also from the previous $L$ steps.
%
We will demonstrate the effectiveness and tune each of the options (PCA / linear approximation as well as lifespan option) in our evaluation.

The new approach is given in Algorithm~\ref{alg:new}.
%
In this algorithm, instead of fixing the set of templates, we compute one set of templates (that is, a collection of $n$ template directions), using linear approximation of the dynamics and PCA.
%
The algorithm makes use of helper function \texttt{hstack}, which converts a bunch of column vectors into a matrix (as shown in Equation~\ref{eq:ptopeexample} provided in Example~\ref{ex:ptope}).
%
The notation $M_{*,i}$ is used to refer to the $i^{th}$ column of matrix $M$.
%
The \texttt{Maximize} function takes in a parallelotope bundle $Q$ and direction vector $v$ (one of the template directions), and returns the point $p \in Q$ that maximizes the dot product $v \cdot p$ (for computing support points).
%
This can be computed efficiently using linear programming.
%
The \texttt{ApproxLinearTrans} function computes the best approximation of a linear transformation given a list of points before and after the one-step transformation $f$.
%
More specifically, given a matrix $X$ of points before applying the transformation $f$, a matrix of points after the transformation $X'$, we perform a least-squares fit for the linear transition matrix $A$ such that $X' \approx AX$.
%
This can be computed by $A = X' X^\dagger$, where $X^\dagger$ is the Moore-Penrose pseudoinverse of $X$.
%
The \texttt{PCA} function returns a set of orthogonal directions using principal component analysis of a set of points.
%
Finally, \texttt{TransformBundle} is the same as in Algorithm~\ref{alg:old}.

@@

Algorithm~\ref{alg:new} computes the dynamic templates for each time step $k$.
%
Line~\ref{ln:linearapprox} computes the linear approximation of the nonlinear dynamics and this linear approximation is used to compute the new template directions according to this linear transformation in Line~\ref{ln:linapp}.
%
The PCA directions of the images of support points is computed in line~\ref{ln:pca}.
%
For the time step $k$, the linear and PCA templates are given as $\T_{k}^{lin}$ and $\T_{k}^{pca}$, respectively.
%
To improve the accuracy of the reachable set, we compute the reachable set with respect to not just the template directions at the current step, but with respect to other template directions for time steps that are within the lifespan $L$.
%
This is because, the nonlinear systems we deal with are often not chaotic in nature.
%
Therefore, the templates at time step $k-1$ are very similar to that of template directions at time step $k$.
%

>>>>

\vspace{-1.2em}
\section{Reachability Algorithm}
\label{sec:theory}

In this work, we develop parallelotope reachability algorithms that are \textbf{automatic} with \textbf{dynamic} parallelotopes.
%
The state of the art, in contrast, is \textbf{manual}, where the user specifies a set of parallelotope directions at the beginning.
%
The parallelotopes are also \textbf{static} and do not change during the course of the computation.
%
In this section, we detail the modifications to the algorithm and present their correctness arguments.

@@

We first present the original algorithm\cite{dang2012reachability} where the user manually specifies the number of parallelotopes and a set of static directions for each parallelotope.

Recall the system is $n$-dimensional with dynamics function $f: \reals^{n} \rightarrow \reals^{n}$.
%
The parallelotope bundle $Q$ is specified as a collection of $m$ template directions $\T^{Q} \in \reals^{m \times n} (m > n)$ and the set of constraints that define each of the member parallelotopes.
%

% consisting of $m$ parallelotopes , the user should manually provide $n$ generator directions for each parallelotope.
%
% We write the collective set of generator directions as $\mathcal{G}$, which is a set of sets, each of which contains $n$ direction vectors.
%
Another input to the algorithm is the initial set, given as a parallelotope $P_0$.
%
When the initial set is a box, $P_0$ consists has axis-aligned template directions.
%
The output of the algorithm is, for each step $k$, the set $\overline\Theta_k$, which is an overapproximation of the reachable set at step $k$, $\Theta_k \subseteq \overline\Theta_k$.

The high-level pseudo-code is written in Algorithm~\ref{alg:old}.
%
The algorithm simply calls \tbundle for each step, producing a new parallelotope bundle computed from the previous step's bundle.
%
To compute the image of $Q$, the algorithm computes the upper and lower bounds of $f(x)$ with respect to each template direction.
%
Since computing the maximum value of $f(x)$ along each template direction on $Q$ is computationally difficult, the algorithm instead computes the maximum value over each of the constituent parallelotopes and uses the minimum of all these maximum values.
%
The \tbundle operation works as follows.
%
Consider a parallelotope $P$ in the bundle $Q$.
%
From the definition, it follows that $Q \subseteq P$.
%
% and half-space representation as $\tup{\T, c_{l}, c_{u}$.
Given a template direction $\T_i$, the maximum value of $\T_{i} f(x)$ for all $x \in Q$ is less than or equal to the maximum value of $\T_{i} f(x)$ for all $x \in P$.
%
Similar argument holds for the minimum value of $T_{i} f(x)$ for all $x \in Q$.
%
% We would like to compute the image of the parallelotope $P$ with respect to the nonlinear transformation $f(x)$.
%
% That is, we would like to compute $P'$ such that $\forall x \in P, f(x) \in P'$.
%
% Given that current techniques use static template directions, the parallelotope $P'$ also uses the template directions $\T^{P}$.
%
To compute the upper and lower bounds of each template direction $\T_{i} f(x)$, for all $x \in P$, we perform the following optimization.
%
\begin{eqnarray}
  \max ~ \T_i^{P} \cdot f(x) \label{eq:maxf}\\
  s.t. ~~ x \in P.\nonumber
\end{eqnarray}

Given that $P$ is a parallelotope, all the states in $P$ can be expressed as a vector summation of anchor and scaled generators.
%
Let  $\tup{a, G}$ be the generator representation of $P$.
%
The optimization problem given in Equation~\ref{eq:maxf} would then transform as follows.

@@

Equation~\ref{eq:maxalpha} is a form of $\mathsf{optBox(\T_{i} \cdot f)}$ over $[0,1]^n$.
%
One can compute an upper-bound to the constrained nonlinear optimization by invoking one of the Bernstein polynomial or interval-arithmetic-based methods.
%
Similarly, we compute the lowerbound of $\T_{i}f(x)$ for all $x \in P$ by computing the upperbound of $-1 \times \T_{i}f(x)$.
%

We iterate this process (i.e., computing the upper and lower bound of $T_{i}f(x)$) for each parallelotope in the bundle $Q$.
%
% Observe that the role of the parallelotopes in the bundle is to compute an upper bound on the maximum value of $T_{i}f(x)$ by expanding the domain from $Q$ to the corresponding parallelotope $P$.
%
Therefore, the tightest upper bound on $T_{i}f(x)$ over $Q$ is the least of the upper bounds computed from each of the parallelotopes.
%
A similar argument holds for lower bounds of $T_{i}f(x)$ over $Q$.
%
Therefore, the image of the bundle $Q$ will be the bundle $Q'$ where the upper and lower bounds for templates directions are obtained by solving several constrained nonlinear optimization problems.

@@

The proposed automatic dynamic algorithm does not require the user to provide the set of template directions $\T$; instead it generates these templates directions automatically at each step.
%
We use two techniques to generate such template directions, first: computing local linear approximations of the dynamics and second, performing principal component analysis (PCA) over sample trajectories.
%
To do this, we first sample a set of points in the parallelotope bundle called \emph{support points} and propagate them to the next step using the dynamics $f$.
%
Support points are a subset of the vertices of the parallelotope that either maximize or minimize the template directions.

Intuitively, linear approximations can provide good approximations when the dynamics function is a time-discretization of a continuous system.
%
In this case, for small time steps a nonlinear function can be approximated fairly accurately by a linear transformation.
%
We use the support points as a data-driven approach to find the best-fit linear function to use.
%
If the dynamics of a system is linear, i.e., $x^{+} = Ax$, the image of the parallelotope $c_{l} \leq \T x \leq c_{u}$, is the set $c_{l} \leq \T \cdot A^{-1} x \leq c_{u}$.
%
Therefore, given the template directions of the initial set as $\T_0$, we compute the local linear approximation of the nonlinear dynamics and change the template directions by multiplying them with the inverse of the approximate linear dynamics.
%
The second technique for generating template directions performs principal component analysis over the images of the support points.
%
Using PCA is a reasonable choice as it produces orthonormal directions that can construct a rotated box for bounding the points.
%

Observe that in general, the dynamics is nonlinear and therefore, the reachable set could be non-convex.
%
On the other hand, a parallelotope bundle is always a convex set.
%
To mitigate this discrepancy, we can improve accuracy of this representation by considering more template directions.
%
For this purpose, we use a notion of \emph{template lifespan}, where we use the linear approximation and/or PCA template directions not only from the current step, but also from the previous $L$ steps.
%
We will demonstrate the effectiveness and tune each of the options (PCA / linear approximation as well as lifespan option) in our evaluation.

The new approach is given in Algorithm~\ref{alg:new}.
%
In this algorithm, instead of fixing the set of templates, we compute one set of templates (that is, a collection of $n$ template directions), using linear approximation of the dynamics and PCA.
%
The algorithm makes use of helper function \texttt{hstack}, which converts column vectors into a matrix (as shown in Equation~\ref{eq:ptopeexample} provided in Example~\ref{ex:ptope}).
%
The notation $M_{*,i}$ is used to refer to the $i^{th}$ column of matrix $M$.
%
The \texttt{Maximize} function takes in a parallelotope bundle $Q$ and direction vector $v$ (one of the template directions), and returns the point $p \in Q$ that maximizes the dot product $v \cdot p$ (for computing support points).
%
This can be computed efficiently using linear programming.
%
The \texttt{ApproxLinearTrans} function computes the best approximation of a linear transformation given a list of points before and after the one-step transformation $f$.
%
More specifically, given a matrix $X$ of points before applying the transformation $f$, a matrix of points after the transformation $X'$, we perform a least-squares fit for the linear transition matrix $A$ such that $X' \approx AX$.
%
This can be computed by $A = X' X^\dagger$, where $X^\dagger$ is the Moore-Penrose pseudoinverse of $X$.
%
The \texttt{PCA} function returns a set of orthogonal directions using principal component analysis of a set of points.
%
Finally, \texttt{TransformBundle} is the same as in Algorithm~\ref{alg:old}.

@@

Algorithm~\ref{alg:new} computes the dynamic templates for each time step $k$.
%
Line~\ref{ln:linearapprox} computes the linear approximation of the nonlinear dynamics and this linear approximation is used to compute the new template directions according to this linear transformation in Line~\ref{ln:linapp}.
%
The PCA directions of the images of support points is computed in line~\ref{ln:pca}.
%
For the time step $k$, the linear and PCA templates are given as $\T_{k}^{lin}$ and $\T_{k}^{pca}$, respectively.
%
To improve the accuracy of the reachable set, we compute the overapproximation of the reachable set with respect to not just the template directions at the current step, but with respect to other template directions for time steps that are within the lifespan $L$.
%
%This is because, the nonlinear systems we deal with are often %not chaotic in nature.
%
%Therefore, the templates at time step $k-1$ are very similar %to that of template directions at time step $k$.
%
